---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# INDEED
## Overview


This package implements INDEED algorithm from Zuo et. al.'s Methods paper: INDEED: Integrated differential expression and differential network analysis of omic data for biomarker discovery (PMID: 27592383). 

This R package will generate a csv file containing information such as p-values, node degree and activity score for each biomolecule. A higher activity score indicates that the corresponding biomolecule has more neighbors connceted in the differential network and their p-values are more statistically significant. It will also generate a csv file for the differential network created by INDEED. 

## Installation

You can install INDEED from github with:

```{r gh-installation, eval=FALSE}
# The development version from GitHub:
# install.packages("devtools")
devtools::install_github("cx30/INDEED")
```

## Usage

knitr::opts_knit$set(progress=FALSE)

```{r example, eval=TRUE, quiet=TRUE}

library(INDEED)

select_sig(Met_GU, Met_Group_GU, Met_name_GU, partial = TRUE)
```


